Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines

### 2013-03-28 paper/paper.tex
Reordering and semantics changes complete. Paper way too long.

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance \emph{and} precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis. The
difference is that the store is not modified in-place, but updated
after all frontier states have been processed. This has implications
for the analysis' precision and determinism. Specifically, higher
precision, and it is deterministic even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

% Notice that now $S$ has several copies of the abstract store in it. As
% it is, this semantics is much less efficient (but still more precise)
% than the previously proposed semantics because membership checks have
% to compare entire stores. We return to this in section
% \ref{sec:timestamping}.

For clarity, we will exhibit non-widened semantics unless a technique
specifically requires it. The final step to the baseline takes the
complexity from quartic to cubic in the monovariant case.

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, preallocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}

@@

The move to the imperative will be made last in order to show the
effectiveness of these techniques in the purely functional realm.

\subsection{Timestamping monotonicity}

The frontier-based semantics of the baseline has a severe bottleneck:
checking if a state has been explored or not. Given two states,
checking equality is expensive because the stores within each are
large, and every entry must be checked against every other. Hashes can
sometimes rule out inequality relatively quickly, but the incidence of
collisions and actual equality is costly.

@@

There are two subtle losses of precision in Shivers' original
time-stamp technique that we can fix.

@@

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective
application positions. 
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.3]{fanout}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). 
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.3]{lazy}
\end{center}

@@

\begin{align*}
\spchoice
{\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
 \force(\mval) &= \{\mval\}}
{\force &: \Value \to \Set(\Value) \\
 \force(\superposition{\mval{s}}) &= \mval{s} \\
 \force(\mval) &= \{\mval\}}
\\
%
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
%
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^\mcntr]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
\text{ where }
\maddr_f &= \alloc(\mstate) \\
\msto' &=
\spchoice
{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
%
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^\mcntr]{\mexpi,\menv,\msto,\mkont} \\
\text{ if } \mkont &\in \msto(\maddr_\mkont),
            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\end{align*}
Since {\tt if} guards are in strict position, we must force the value
to determine which branch to to take. The middle rule uses $\force$
only to combine with values in the store - it does not introduce
needless non-determinism.

@@

The following example illustrates
the essence of abstract compilation effect:
\[
\mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
\]
makes the following transitions:
\begin{align}
& \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
\machstep\; &
\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_{f1},\maddr_{a1},\maddr_1}}
\\
\machstep\; &
\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_{f2},\maddr_{a2},\maddr_2}}
\\
\machstep\; &
\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3}} % {\mexp_2}
\\
\machstep\; &
\spchoice
{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
\end{align}
where \(
\msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
\maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
\maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).

Whereas abstract compilation gives us in one step:
\begin{align*}
\compile{\mexp}^\mcntr(\menv,\msto_0,\mkont) = 
\spchoice
{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\saddr{\menv(\mvar)},\msto'_4}}
{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
\end{align*}
where \(
\msto'_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
[\maddr_2 \mapsto \kar{\compile{\mexp_3},\menv,\maddr_1}],
[\maddr_3 \mapsto \kar{\compile{\mexp_2},\menv,\maddr_2}]\).

@@

\begin{figure}
\begin{align*}
\compile{\_} &: (\Expr \to \Env \times \Store \times \Kont \times \Counter) \to \State \\
\mcntr' &= \tick(\mlab,\menv,\msto,\mcntr) \\
\compile{\svar\mvar} &=
 \lambda^\mcntr(\menv,\msto,\mkont) .
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}}
\\
\compile{\slit\mlit} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile{\mexp},\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
\\
&\setlength\arraycolsep{5pt}
\begin{array}{lrl}
\text{ where } & \maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
               & \msto' = \msto\sqcup[\maddr_\mkont \mapsto \setof{\mkont}]
\end{array}
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr_\mkont \mapsto \setof{\mkont}]
\end{align*}
\caption{Abstract compilation}
\label{fig:compile}
\end{figure}

>>>>

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \widehat{\State} \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, preallocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}

@@

We apply the frontier-based semantics combined with timestamped stores
as our first step.  The move to the imperative will be made last in
order to show the effectiveness of these techniques in the purely
functional realm.

\subsection{Timestamped frontier}

The semantics given for store widening in section \ref{sec:storewiden},
while simple, is wasteful. It also does not model what typical
implementations do. It causes all states found so far to step each
iteration, even if they are not revisited. This has negative
performance \emph{and} precision consequences (changes to the store
can travel back in time in straight-line code). We instead use a
frontier-based semantics that corresponds to the classic ``worklist''
algorithms for analysis. The difference is that the store is not
modified in-place, but updated after all frontier states have been
processed. This has implications for the analysis' precision and
determinism. Specifically, higher precision, and it is deterministic
even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } I &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              F' &= \{c \mid (c,\_) \in I, (c,\msto') \notin S\} \\
              S' &= \setof{(c,\msto') \mid c \in F'} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

Notice that now $S$ has several copies of the abstract store in it. As
it is, this semantics is much less efficient (but still more precise)
than the previously proposed semantics because membership checks have
to compare entire stores. Checking equality is expensive because the stores within each are
large, and every entry must be checked against every other. Hashes can
sometimes rule out inequality relatively quickly, but the incidence of
collisions and actual equality is costly.

@@

There are two subtle losses of precision in Shivers' original
timestamp technique that we can fix.

@@

\begin{align*}
\Sigma &\in \Store^* \\
S &\subseteq {\mathbb N} \times \widehat{\State} \\
F &\subseteq \widehat{\State} \\
(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              (t',\Sigma') &=\left\{\begin{array}{ll}
                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
                           (t,\Sigma)   & \text{ otherwise}
                          \end{array}\right. \\
              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
              S' &= \setof{(c,t') \mid c \in F'}\\
\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

The observation Shivers made was that the store is increasing
monotonically, so all stores throughout execution will be totally
ordered (form a chain). This observation allows you to replace stores
with pointers into this chain. We keep the stores around in $\Sigma$
to acheive a complete abstraction. This corresponds to the temporal
information about the execution's effect on the store.

Note also that $F$ is only populated with states that have not been
seen at the resulting store. This is what produces the more precise
abstraction than the baseline widening.

The general fixed-point combinator we showed above can be specialized
to this semantics, as well. In fact, $\widehat{\machstep}^T$ is a functional
relation, so we can get the least fixed-point of it directly.

\begin{lemma}
  $\widehat{\machstep}$ maintains the invariant that all stores in $S$ are
  totally ordered and $\msto$ is an upper bound of the stores in $S$.
\end{lemma}

\begin{lemma}
  $\widehat{\machstep}^T$ maintains the invariant that $\Sigma$ is in
  order with respect to $\sqsupset$ and $\msto = \hd(\Sigma)$.
\end{lemma}

\begin{theorem}
Timestamps are a complete abstraction
\end{theorem}
The proof follows from the galois connection that, in one direction,
sorts all the stores in $S$ to form $\Sigma$, and translates stores in
$S$ to their distance from the end of $\Sigma$. In the other
direction, timestamps are replaced by the stores they point to.

The above technique requires joining entire (large) stores
together. Additionally, there is still a comparison of stores, which
we established is expensive. Not every step will modify all addresses
of the store, so joining entire stores is wasteful in terms of memory
and time. We can instead log store changes and replay the change log
on the full store after all steps have completed, noting when there is
an actual change. This uses far fewer join and comparison operations,
leading to less overhead, and is precision-preserving.

We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
 becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:
\begin{equation*}
\replay : (\StoreDelta \times \Store) \to (\Store \times \Boolean)
\end{equation*}
\begin{align*}
\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &=
\ttuple{\msto'}
       {\mval{s_i} \overset{?}{=} \msto'(\maddr_i) \vee \ldots} \\
\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots}
\end{align*}

We change the semantics slightly to add to the change log rather than
produce an entire modified store.  The transition relation is
identical except for the addition of this change log.  We maintain the
invariant that lookups will never rely on the change log, so we can
use the originally supplied store unmodified.

A taste of the changes to the reduction relation is as follows:

\begin{align*}
\dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \dmachstep_\msto
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
\text{ where }\maddr &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr]
\end{align*}

% Compilation changes to additionally take a $\msdiff$ component, so the
% above rule's right hand side would instead be $k^{\mcntr'}(\menv',
% \msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.

We lift $\dmachstep$ to accommodate for the asymmetry
in the input and output, and change the frontier-based semantics in the following way:

\begin{align*}
(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(\mastate',\msdiff) \mid \mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{\mastate \mid (\mastate,\_) \in I, (\mastate,t') \notin S} \\
 S' &= \setof{(\mastate,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}

\begin{lemma}
$\mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff}$ iff
$\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
\end{lemma}
By cases on $\dmachstep$ and $\machstep$.

\begin{lemma}[$\changep$ means change]
Let $\replay(\msdiff,\msto) = \ttuple{\msto'}{\changep}$. $\msto' \neq \msto$ iff $\changep$.
\end{lemma}
By induction on $\msdiff$.

\begin{theorem}
$\damachstep$ is a complete abstraction of $\widehat{\machstep}^T$.
\end{theorem}
Follows from previous lemma and that join is associative and commutative.

% Correctness here requires a lemma that relates the differences in the compilation functions.

% \begin{lemma}[Compile coherence]
% Let $\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$
% and $\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \wn(\mastate',\msto^{*'})$.
% If $\msto \equiv \msto^*$, $\msdiff \equiv \msdiff^*$, and $\mkont \equiv \mkont^*$, then
% $\mastate' \equiv \mastate$
% and there exists an $\msdiff''$ such that \\
%    $\replay(\msdiff'',\replay(\msdiff^*,\msto^*)) \equiv \replay(\msdiff',\msto)$.
% \end{lemma}
% Here $\equiv$ means structurally the same, only compiled expressions
% are compiled with the updated compilation function that builds store
% deltas. Proof is by induction on $\mexp$.

% \begin{theorem}[Equivalence with wide abstract compilation]
% If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
% $(S,F,\msto) \machstep (S',F',\msto')$ iff
% $\exists S^*_1, F^*_1, \msto^*_1.
%   S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
%   (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
% \end{theorem}
% Proof by the above lemma and that replaying in any order is equivalent
% (associativity and commutativity of $\sqcup$).

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective
application positions.
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.3]{fanout}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if}, a called procedure, or a numerical primitive application).
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.3]{lazy}
\end{center}

@@

\begin{align*}
\spchoice
{\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
 \force(\msto,\mval) &= \setof{\mval}}
{\force &: \Value \to \Set(\Value) \\
 \force(\superposition{\mval{s}}) &= \mval{s} \\
 \force(\mval) &= \{\mval\}}
\\
%
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
%
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
\text{ where }
\maddr_f &= \alloc(\mstate) \\
\msto' &=
\spchoice
{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
%
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
\text{ if } \mkont &\in \msto(\maddr_\mkont),
            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\end{align*}
Since {\tt if} guards are in strict position, we must force the value
to determine which branch to to take. The middle rule uses $\force$
only to combine with values in the store - it does not introduce
needless non-determinism.

@@

The following example illustrates
the essence of abstract compilation effect:
\[
\mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
\]
makes the following transitions:
\begin{align}
& \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
\machstep\; &
\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_1}}
\\
\machstep\; &
\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_2}}
\\
\machstep\; &
\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_3}} % {\mexp_2}
\\
\machstep\; &
\spchoice
{\co{\kar{\mexp_1,\menv,\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
\end{align}
where \(
\msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
\maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
\maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).

Whereas abstract compilation gives us in one step:
\begin{align*}
\compile{\mexp}^\mcntr(\msto_0)(\menv,\epsilon,\mkont) =
\spchoice
{\co{\kar{\compile{\mexp_1},\menv,\maddr_3},\saddr{\menv(\mvar)}},\msdiff_4}
{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
\end{align*}
where
\begin{equation*}
\msdiff_4 = [\ttuple{\maddr_1}{\setof{\mkont}},
             \ttuple{\maddr_2}{\setof{\kar{\compile{\mexp_3},\menv,\maddr_1}}},
             \ttuple{\maddr_3}{\setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}}]\text.
\end{equation*}

@@

\begin{figure}
\begin{align*}
\compile{\_} &: \Expr \to \Store \\
             &\phantom{: \Expr } \to \Env  \times \StoreDelta \times \Kont \times \Counter \\
             &\phantom{: \Expr } \to \State \\
\mcntr' &= \tick(\mlab,\menv,\msto,\mcntr) \\
\compile{\svar\mvar}_\msto &=
 \lambda^\mcntr(\menv,\msdiff,\mkont) .
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)}},\msdiff}
{\co{\mkont,\superposition{\msto(\menv(\mvar))}},\msdiff}
\\
\compile{\slit\mlit}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
\co{\mkont,\mlit},\msdiff
\\
\compile{\slam\mvar\mexp}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
\co{\mkont,\clos{\mvar,\compile{\mexp},\menv}},\msdiff
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
\\
&\setlength\arraycolsep{5pt}
\begin{array}{lrl}
\text{ where } & \maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
               & \msdiff' = \cons{\ttuple{\maddr_\mkont}{\setof{\mkont}}}{\msdiff}
\end{array}
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&\phantom{\text{ where }} \msdiff' = \cons{\ttuple{\maddr_\mkont}{\setof{\mkont}}}{\msdiff}
\end{align*}
\caption{Abstract compilation}
\label{fig:compile}
\end{figure}

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \setof{ \inject(\compile{\mexp}^\mtcntr_\bot(\bot,\epsilon,\kmt)) \multimachstep \mstate}
                    \text{ where } \\
\inject(\mastate,\msdiff) &= \wn(\mastate,\replay(\msdiff,\bot)) \\
\wn(\mastate,\msto) \machstep \wn(\mastate',\msto') &\iff \mastate \cmachstep_\msto \mastate',\msdiff \\
\msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
\end{align*}
\\[2mm]
\begin{align*}
%% CONTINUE
\co{\kmt,\mval} &\cmachstep_\msto
\ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_\mkont},\mval} & \cmachstep_\msto
\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}) \\
\text{ where } \maddr_f &= \alloc(\mstate) \\
               \msdiff &= \cons{\ttuple{\maddr_f}{\force(\msto,\mval)}}{\epsilon}
\\
\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont},\mval} & \cmachstep_\msto
\ap[^\mcntr_\mlab]{\mval,\mval\mkont},\epsilon \\
\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue} & \cmachstep_\msto
\mcompi{0}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse} & \cmachstep_\msto
\mcompi{1}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\mkont} & \cmachstep_\msto
\mcomp^{\mcntr'}(\msto)(\menv',\msdiff,\mkont) \\
\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
              \msdiff &= \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon}
\\
\ap{\mop,\mval,\mkont} & \cmachstep_\msto
\co{\mkont,\mval'},\epsilon \\
\text{ where } \mvalx{u} &\in \spchoice{\force(\msto,\mval)}{\force(\mval)}, \mval'\in\interpdelta(\mop,\mvalx{u})
\end{align*}
\end{gather*}
\caption{Abstract abstract machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

\paragraph{Correctness}
The correctness of abstract compilation seems obvious, but it has
never before been rigorously proved. What constitutes correctness in
the case of dropped states, anyway? Applying an abstract bytecode's
function does many ``steps'' in one go, at the end of which, the two
semantics line up again (modulo representation of expressions). This
constitutes the use of a notion of stuttering. We provide a formal
analysis of abstract compilation \emph{without} store widening with a proof
of a stuttering bisimulation~\cite{ianjohnson:BCG88} between this
semantics and lazy non-determinism without widening to show precision
preservation. We revisit store widening at the end of the section.
